Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-12 paper/paper/SPIE/article.tex
[WIP]

\newcommand{\code}[1]{\small \texttt{#1} \normalsize}
\newcommand{\tcode}[1]{\footnotesize \texttt{#1} \normalsize}

@@

\section*{Results}
% df.groupby(["subgroup", "gross_feature"])["auroc"].median()
Across all classification tasks and analytic choices, we find eiegenfeatures to
more often have predictive utility than not (82.4\% of median AUROCs \(>\) 0.5;
median AUROC range across classification tasks 0.47 - 0.64). Simple baseline
reductions on source timeseries, by contrast, were less useful (58.8\% of
median AUROCs \(>\) 0.5, median AUROC range across classification tasks 0.42 -
0.62). Additionally, eigenfeature AUROC distributions were overall much more
right-tailed than baseline features suggesting greater predictive potential.
However, performance distributions were wide, and often significantly affected by
analytic choices.

@@

At its most rudimentary, RMT describes the expected behaviour of the
eigenvalues—also often called levels—of a number of classes of random
matrices\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}. A random matrix is,
simplifying somewhat, a matrix in which the entries are drawn from
independently and identically distributed (\textit{iid}) distributions. Though
first developed to describe the fluctuation of nuclei energy levels in quantum
physics \cite{mehtaRandomMatrices2004, guhrRandommatrixTheoriesQuantum1998a},
RMT has been shown have extremely broad potential.

In small-scale physical systems, RMT universalities have been observed in
quantum chaotic systems, complex nuclei, atoms, molecules and disordered
mesoscopic systems
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001},
and at larger scales, RMT has been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002} , social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids
\cite{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009}
and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}.
Within biological systems, RMT has also been used to successfully model aspects
of  amino acid functional relationships
\cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in epileptic seizures
\cite{osorioPhasesynchronizationRandommatrixBased2011}, and in protein-protein
interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
even more recently, has provided insights into the behaviours of deep
neural networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

@@

For an imaging modality like fMRI, where changes in the
blood-oxygenation-level-dependent (BOLD) signals are related to neural
activity, RMT may be an ideal starting point. Each voxel time course (or
collection of such time sources - i.e. ROIs) can be considered an interacting
component of the system. Likewise, in \textit{functional connectivity}
analyses, statistical relationships of the BOLD ROI time-courses are
investigated in the hope of gaining insights into brain functioniong (see
\citenum{bastosTutorialReviewFunctional2016,
vandenheuvelExploringBrainNetwork2010} for reviews, but also
\citenum{nobleDecadeTestretestReliability2019} and
\citenum{lurieQuestionsControversiesStudy2020} for challenges facing functional
connectivity analyses). In this framework, each connection or correlation can
be considered a system component, and the eigenvalues of such a correlation
matrix can be examined from the perspective of RMT.

@@

RMT has also been used in ROI-based fMRI functional-connectivity studies to
investigate differences between rest and task states
\cite{wangSpectralPropertiesTemporal2015}, between subjects with and without
ADHD \cite{wangRandomMatrixTheory2016}, and between pain and non-pain states
\cite{matharooSpontaneousBackpainAlters2020}. Across these three studies, the
spectra of resting or low-attention states exhibited properties close to the
GOE. These findings suggest that certain aspects of psychological processes
might be characterized, in part, by features computed from the eigenvalues of
fMRI correlation matrices, and that these features might vary in an
interpretable manner across psychological processes. If this is the case, RMT
could aid in understanding the functioning of the human brain.

\subsection{Eigenvalues and Prediction}

The basic insight of RMT is thus that the eigenvalues alone may provide
interesting information about highly complex systems. However, real systems are
usually noisy, and involve a mixture of random and non-random components and
interactions. RMT is statistical in nature, describing only the expected
behaviour of the spectra of \textit{iid} random matrics. To this end, a number
of summary statistics or ``spectral observables''
\cite{mehtaRandomMatrices2004, guhrRandommatrixTheoriesQuantum1998a} can be
computed from empirically observed spectra, with these spectral observables
sometimes being better suited for further analysis, or for the comparison of
empirical observations to theory. Two such summary statistics that have been
popular\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014, matharooSpontaneousBackpainAlters2020,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016} are the \textit{spectral rigidity} and
\textit{level number variance} (``rigidity'' and ``level variance'' for short;
details in Section \ref{sec:methods}).

However, from a predictive standpoint, these summary statistics may eliminate
predictively-useful information. If the basic RMT insight is that the
eigenvalues alone can provide understanding of a system, then those eigenvalues
(or other simple transformations of them) ought also to be predictively
useful. We examine a number of such \textit{eigenfeatures} in this study.

In addition, the functional connectivity - the matrix of correlations of
various collections of voxels of an fMRI scan - is a priori a useful
representation of the fMRI data. However, the full correlation matrix between
all voxels is itself often computationally infeasible to work with, and thus is
reduced in various ways prior to being used in analyses.

@@

RMT suggests a potentially useful reduction in the form of the eigenvalues of
the voxelwise functional connectivity. An \(N \times t\) matrix \(\bf{A}\) of
\(N\) time series of length \(t\), and where \(t \ll N\) will have a symmetric
correlation (or covariance) matrix with \(t - 1\) non-zero, positive
real-valued eigenvalues. These eigenvalues can be computed highly
efficiently via transposition of the voxelwise correlation or covariance
matrix.

@@

Previous studies applying RMT to functional connectivity
data\cite{wangRandomMatrixTheory2016,
wangSpectralPropertiesTemporal2015a,matharooSpontaneousBackpainAlters2020,
guRandomMatrixTheory2020} took largely descriptive / explanatory approaches,
noting only differences in RMT across subgroups and/or conditions. For example
\citenum{wangRandomMatrixTheory2016,guRandomMatrixTheory2020} use a
Kolmogorov-Smirnov test, and \citenum{wangSpectralPropertiesTemporal2015a} uses
a t-test to provide evidence of differences in various RMT metrics. However,
statistically significant differences in a metric do not imply practical
significance or predictive utility \cite{loWhySignificantVariables2015}.

In addition, the previous papers do not provide publicly-accessible code to
reproduce results. The extraction of the spectral rigidity and level variance
is computationally-demanding and non-trivial, requiring an \textit{unfolding}
procedure\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004},
which is, unfortunately, a poorly-documented fitting procedure that requires
making a number of arguably subjective decisions. These fitting decisions can
dramatically impact RMT conclusions
\cite{abul-magdUnfoldingSpectrumChaotic2014,
abueleninSpectralUnfoldingChaotic2018, fossionRandommatrixSpectraTime2013,
abueleninEffectUnfoldingSpectral2012, moralesImprovedUnfoldingDetrending2011}.
On top of the flexibility in the implementation and application of RMT to
empirical data, there is also the flexibility introduced by the choices in the
complex preprocessing pipelines of fMRI\cite{parkerBenefitSliceTiming2019}.
These "researcher degrees of
freedom"\cite{simmonsFalsePositivePsychologyUndisclosed2011b} in combination
with the absence of clear code, and exclusively descriptive/explanatory
analytic approaches means that the practical utility of RMT for fMRI is largely
uninvestigated.

@@

We selected fMRI data publicly available on the
\href{https://openneuro.org/}{OpenNeuro} platform (36). Selection criteria was
somewhat subjective, but we required (1), that all scans have the same spatial
and temporal resolutions, (2), the the dataset comprise 10+ human subjects, and
(3), that it was reasonable to classify at the level of an entire scan, i.e.
neither task nor event timings were needed to determine class membership. This
yielded seven datasets.

@@

\begin{table}[h!]
\caption{
    ID = Identifier for paper. Subgroup = name of subgroup. Subjects = Number of subjects.
    Task = fMRI task. ANT = Attention Network Task \cite{fanActivationAttentionalNetworks2005}
}
\label{table:2}
\small
\centering
\begin{tabular}{ l r c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
Aging   &  older             & 28      & resting-state        & 1 \\
        &  younger           & 34      & resting-state        & 1 \\
        &                    &         &                      &   \\
Bilingual  &  bilingual          & 59      & resting-state    & 1 \\
        &  monolingual        & 33      & resting-state       & 1 \\
        &                    &         &                      &   \\
Depress &  depression         & 51      & resting-state       & 1 \\
        &  control            & 21      & resting-state       & 1 \\
        &                    &         &                      &   \\
Learn   &  task              & 24      & learn image sketches & 16\\
        &  rest              & 24      & resting-state        & 2 \\
        &                    &         &                      &   \\
Osteo   &  duloxetine        & 19      & resting-state        & 1 \\
        &  pain              & 37      & resting-state        & 1 \\
        &  nopain            & 20      & resting-state        & 1 \\
        &                    &         &                      &   \\
Park    &  Parkinson's       & 25      & ANT                  & 12\\
        &  control           & 21      & ANT                  & 12\\
        &                    &         &                      &   \\
Attention&  vigilant           & 11     & resting-state       & 2 \\
        &  nonigilant         & 11     & resting-state        & 2 \\
        &  trait-attentive     & 11    & resting-state        & 2 \\
        &  trait-non-attentive & 11    & resting-state        & 2 \\
        &  task-attentive     & 11     & resting-state        & 2 \\
        &  task-nonattentive  & 11     & resting-state        & 2 \\
\hline
\end{tabular}
\end{table}

@@

The \textit{Depression} dataset\cite{bezmaternykhddRestingStateClosed2021,
bezmaternykhBrainNetworksConnectivity2021} includes rs-fMRI scans from
non-depressed controls and mildy- or moderately-depressed subjects. In our analysis,
the binary classification task is \code{depress v control}.

@@

The \code{viglant v nonvigilant} task was formed based on self-report
questionnaire items involving
"vigilance"\cite{gorgolewskiCorrespondenceIndividualDifferences2014}. The
\code{trait\_attend v trait\_nonattend} task was formed using PANAS-X
\cite{watsonPANASXManualPositive1994} items related to self-reported
wakefulness and attention over the past weeks. Finally, the
\code{task\_attend v tas\_nonattend} classification task was constructed
using the scores of The Conjunctive Continuous Performance Task
(CCPT\cite{shalevConjunctiveContinuousPerformance2011}). This is a behavioural
performance task which requires the subject to quickly and selectively respond
to only certain visual stimuli. Details allowing exact reproduction of these
splits are available in this paper's
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/preprocess/unify\_attention\_data.py}{source
code}.

@@

We limited preprocessing to a simple pipeline of, in order, (1) brain
extraction, (2) slice timing correction, (3), motion correction, and (4),
registration to the MNI152
\cite{fonovUnbiasedAverageAgeappropriate2011,fonovUnbiasedNonlinearAverage2009}
2mm isotropic, asymmetric template version C made available through
\href{https://www.templateflow.org/}{TemplateFlow}
\cite{ciricTemplateFlowFAIRsharingMultiscale2021}.

We limited preprocessing methods to these steps sometimes because data required
for other typical preprocessing steps (e.g. field intensity maps, physiological
measurements) was missing, but also because all pipelined intermediates were saved to later
compare the effect of increasing degrees of pre-processing (see Section
\ref{sec:multiverse}), and it was important to limit the number of
preprocessing degrees to prevent excessive computational costs.

@@

@@

@@

The time and space complexities of computing the full \(N \times N\)
correlation matrix of \(\mathbf{M}\), and the subsequent eigenvalues, are too
large to be tractable. However, since transposition does not change the
eigenvalues, we can use the transpose to efficiently compute these eigenvalues
(see Appendix \ref{sec:transpose}). These raw eigenvalues (feature \code{eigs}
in Table \ref{tab:features}), most directly test the idea that eigenvalues
alone are useful functional connectivity features.

@@

Both the spectral rigidity and level number variance are also computed and
investigated in this study, for all \(L \in \{ 1, 2, \dots 20 \}\).
Because the computation of these statistics is non-trivial, we developed and
release the separate, open-source Python library
\code{empyricalRMT}\cite{dm-bergerStfxecutablesEmpyricalRMTV12022}. The library
allows for efficient, parallel computation of the spectral rigidity and level
variance via Monte Carlo methods, and automatically ensures convergence of the
statistics to user-specified tolerances.

\textit{Trimming Variants}. As the unfolding procedure operates on the sorted
eigenvalues, and involves fitting a smooth polynomial or exponential to these
values, extreme values are often trimmed (omitted from fitting)
\cite{moralesImprovedUnfoldingDetrending2011,
abueleninEffectUnfoldingSpectral2012}. The number of eigenvalues to trim is
highly subjective, and not usually well-documented, and unfortunately, with a
large number of different source matrices, it is too destructive to naively use
the same hard criterion (e.g. largest 10\%, largest 3) for all unfoldings.

@@

\textit{Unfolding Variants}. Following any trimming, the eigenvalues are fit
with a polynomial. The choice of degree has been known to dramatically impact
certain analyses involving the spectral rigidity or level
variance\cite{abul-magdUnfoldingSpectrumChaotic2014,
moralesImprovedUnfoldingDetrending2011, abueleninSpectralUnfoldingChaotic2018,
fossionRandommatrixSpectraTime2013, abueleninEffectUnfoldingSpectral2012}. We
thus compute the rigidity and level variance with all possible combinations of
trimming and unfolding polynomial degrees of 3, 5, 7, and 9.

@@

When computing RMT eigenfeatures, the polynomial fitting during unfolding has a
smoothing effect, as do the additional averaging \(\langle\;\rangle_c\)
operations. However, it is possible other, simpler transformations of the
eigenvalues (e.g. uniform or Savitsky-Golay smoothing) might have equal or
greater predictive utility. We thus also test smoothed variants of the
eigenvalues as predictive features: the sorted eigenvalues are smoothed with
either a uniform (moving average) filter or Savitsky-Golay, using window sizes
of 3, 5, 7, and 9 to yield 8 total ``smoothed eigenvalue'' features.

@@

A combination of RMT and non-RMT eigenfeatures
could be more predictive than either alone. We test this with  a variety of eigenfeature
combinations. The particular combinations are especially chosen to
test whether RMT eigenfeatures add anything beyond smoothing.

Combined features are formed by concatenation, so that if we have features
\(f_1, \dots,  f_n\) with dimensions \(p_1, \dots, p_n\), then the combined
feature is \([f_1; \dots; f_n]\) with \(\sum p_i\) dimensions.  The final
combinations chosen can be found in Table \ref{tab:features}.

>>>>

\newcommand{\code}[1]{\small\texttt{#1}\normalsize}
\newcommand{\tcode}[1]{\footnotesize\texttt{#1}\normalsize}

@@

\section*{Results}
Across all classification tasks and analytic choices, we find RMT- and
eigenvalue-based "eigenfeatures" to have predictive utility more often than
not (82.4\% of median AUROCs \(>\) 0.5; median AUROC range across
classification tasks 0.47 - 0.64). Simple baseline reductions on source
timeseries, by contrast, were less useful (58.8\% of median AUROCs \(>\) 0.5,
median AUROC range across classification tasks 0.42 - 0.62). Additionally,
eigenfeature AUROC distributions were overall more right-tailed than
baseline features, suggesting greater predictive potential. However, performance
distributions were wide, and often significantly affected by analytic choices.

@@

At its most rudimentary, RMT describes the expected behavior of the
eigenvalues—also often called levels—of a number of classes of random
matrices\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}. A random matrix is,
simplifying somewhat, a matrix in which the entries are drawn from
independently and identically distributed (\textit{iid}) distributions. Though
first developed to describe the fluctuation of nuclei energy levels in quantum
physics \cite{mehtaRandomMatrices2004, guhrRandommatrixTheoriesQuantum1998a},
RMT has been shown have extremely broad potential.

In small-scale physical systems, RMT universalities have been observed in
quantum chaotic systems, complex nuclei, atoms, molecules and disordered
mesoscopic systems
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001},
and at larger scales, RMT has been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002} , social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids
\cite{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009}
and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}.
Within biological systems, RMT has also been used to successfully model aspects
of  amino acid functional relationships
\cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in epileptic seizures
\cite{osorioPhasesynchronizationRandommatrixBased2011}, and in protein-protein
interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
even more recently, has provided insights into the behaviors of deep
neural networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

@@

For an imaging modality like fMRI, where changes in the
blood-oxygenation-level-dependent (BOLD) signals are related to neural
activity, RMT may be an ideal starting point, as each voxel time course (or
collection of such time sources - i.e. ROIs) can be considered an interacting
component of the system. Likewise, in functional connectivity
analyses, statistical relationships of the BOLD ROI time-courses are
investigated in the hope of gaining insights into brain functioning (see
\citenum{bastosTutorialReviewFunctional2016,
vandenheuvelExploringBrainNetwork2010} for reviews, but also
\citenum{nobleDecadeTestretestReliability2019} and
\citenum{lurieQuestionsControversiesStudy2020} for challenges facing functional
connectivity analyses). In this framework, each connection or correlation can
be considered a system component, and the eigenvalues of such a correlation
matrix can be examined from the perspective of RMT.

@@

RMT has also been used in ROI-based fMRI functional connectivity studies to
investigate differences between rest and task states
\cite{wangSpectralPropertiesTemporal2015}, between subjects with and without
attention-deficit hyperactive disorder (ADHD)
\cite{wangRandomMatrixTheory2016}, between pain and non-pain states
\cite{matharooSpontaneousBackpainAlters2020}, and between left-sided vs.
right-sided motor imagery \cite{guRandomMatrixTheory2020}. While no differences
were found in the latter study, across the first three studies, the spectra of
resting or low-attention states exhibited properties close to the GOE. These
findings suggest that certain aspects of psychological processes might be
characterized, in part, by features computed from the eigenvalues of fMRI
correlation matrices, and that these features might vary in an interpretable
manner across psychological processes. If this is the case, RMT could aid in
understanding the functioning of the human brain.

\subsection{Eigenvalue Features}

The basic insight of RMT is thus that eigenvalues alone may provide
interesting information about highly complex systems. However, real systems are
usually noisy, and involve a mixture of random and non-random components and
interactions. RMT is statistical in nature, describing only the expected
behavior of the spectra of \textit{iid} random matrices. To this end, a number
of summary statistics or ``spectral observables''
\cite{mehtaRandomMatrices2004, guhrRandommatrixTheoriesQuantum1998a} can be
computed from empirically observed spectra, with these spectral observables
sometimes being better suited for further analysis, or for the comparison of
empirical observations to theory. Two such summary statistics that have been
popular\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014, matharooSpontaneousBackpainAlters2020,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016} are the \textit{spectral rigidity} and
\textit{level number variance} (``rigidity'' and ``level variance'' for short;
details in Section \ref{sec:methods}).

However, from a predictive standpoint, these summary statistics may eliminate
predictively-useful information. If the basic RMT insight is that the
eigenvalues alone can provide understanding of a system, then those eigenvalues
(or other simple transformations of them) ought also to be predictively useful.
We examine a number of such features in this study, and refer to both
RMT-derived features like the rigidity and level variance, and non-RMT-derived
features collectively as \textit{eigenfeatures} in this study.

\subsection{Functional Connectivity Reduction}

The functional connectivity—the matrix of correlations of
various collections of voxels of an fMRI scan—is \textit{a priori} a useful
representation of the fMRI data. However, the full correlation matrix between
all voxels is itself often computationally infeasible to work with, and thus is
reduced in various ways prior to being used in analyses.

@@

RMT suggests a potentially useful reduction in the form of the eigenvalues of
the voxelwise functional connectivity. An \(N \times t\) matrix of
\(N\) time series of length \(t\), and where \(t \ll N\) will have a symmetric
correlation (or covariance) matrix with \(t - 1\) non-zero, positive
real-valued eigenvalues. These eigenvalues can be computed highly
efficiently via transposition of the voxelwise correlation or covariance
matrix.

@@

Previous studies applying RMT to functional connectivity
data\cite{wangRandomMatrixTheory2016,
wangSpectralPropertiesTemporal2015a,matharooSpontaneousBackpainAlters2020,
guRandomMatrixTheory2020} took largely descriptive / explanatory approaches,
noting only differences in RMT across subgroups and/or conditions. For example
\citenum{wangRandomMatrixTheory2016} and \citenum{guRandomMatrixTheory2020} use
a Kolmogorov-Smirnov test, and \citenum{wangSpectralPropertiesTemporal2015a}
uses a t-test to provide evidence of differences in various RMT metrics.
However, statistically significant differences in a metric do not imply
practical significance or predictive utility
\cite{loWhySignificantVariables2015}, nor do they imply generalization or
replicability \cite{amrheinEarthFlat052017}. Cross-validation, by contrast,
attempts to more directly assess these properties
\cite{yarkoniChoosingPredictionExplanation2017}.

In addition, the previous papers do not provide publicly-accessible code to
reproduce results. The extraction of the spectral rigidity and level variance
is computationally-demanding and mathematically and algorithmically
non-trivial, and additionally requires an \textit{unfolding}
procedure\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}.
Unfolding is an exponential fitting procedure that involves a number of highly
subjective decisions regarding outliers and the flexibility of the fitting
function. These decisions are, unfortunately, often poorly or even entirely
undocumented, but are known to often dramatically impact RMT conclusions
\cite{abul-magdUnfoldingSpectrumChaotic2014,
abueleninSpectralUnfoldingChaotic2018, fossionRandommatrixSpectraTime2013,
abueleninEffectUnfoldingSpectral2012, moralesImprovedUnfoldingDetrending2011}.

On top of the flexibility in the implementation and application of RMT to
empirical data, there is also the flexibility introduced by the choices in the
complex preprocessing pipelines of fMRI\cite{parkerBenefitSliceTiming2019}.
These "researcher degrees of
freedom"\cite{simmonsFalsePositivePsychologyUndisclosed2011b} in combination
with the absence of reproducible code and data availability, and predominance
of descriptive and explanatory approaches, raise doubts about the basic robustness
and practical utility of RMT-based functional connectivity analyses.

@@

We selected fMRI data publicly available on the
\href{https://openneuro.org/}{OpenNeuro} platform (36). Selection criteria was
somewhat subjective, but we required (1), the dataset comprise 10+ human
subjects, (2) that all fMRI scans have the same spatial and temporal
resolutions, (3) that the fMRI images could be split into various classes for a
number of binary classification tasks, and (4), that it was reasonable to
classify each scan in its entirety, (i.e. neither task nor event timings
were needed to determine class membership, nor did any scan volumes need to be removed
to make the scan classifiable). This yielded seven datasets total.

@@

\begin{table}[h!]
\caption{
    ID = Identifier for paper. Subgroup = name of subgroup. Subjects = Number of subjects.
    Task = fMRI task. ANT = Attention Network Task \cite{fanActivationAttentionalNetworks2005}
}
\label{table:2}
\small
\centering
\begin{tabular}{ l r c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
Aging   &  older             & 28      & resting-state        & 1 \\
        &  younger           & 34      & resting-state        & 1 \\
        &                    &         &                      &   \\
Bilingual  &  bilingual          & 59      & resting-state    & 1 \\
        &  monolingual        & 33      & resting-state       & 1 \\
        &                    &         &                      &   \\
Depress &  depression         & 51      & resting-state       & 1 \\
        &  control            & 21      & resting-state       & 1 \\
        &                    &         &                      &   \\
Learn   &  task              & 24      & learn image sketches & 16\\
        &  rest              & 24      & resting-state        & 2 \\
        &                    &         &                      &   \\
Osteo   &  duloxetine        & 19      & resting-state        & 1 \\
        &  pain              & 37      & resting-state        & 1 \\
        &  nopain            & 20      & resting-state        & 1 \\
        &                    &         &                      &   \\
Park    &  Parkinson's       & 25      & ANT                  & 12\\
        &  control           & 21      & ANT                  & 12\\
        &                    &         &                      &   \\
Attention&  vigilant           & 11     & resting-state       & 2 \\
        &  nonvigilant         & 11     & resting-state        & 2 \\
        &  trait-attentive     & 11    & resting-state        & 2 \\
        &  trait-non-attentive & 11    & resting-state        & 2 \\
        &  task-attentive     & 11     & resting-state        & 2 \\
        &  task-nonattentive  & 11     & resting-state        & 2 \\
\hline
\end{tabular}
\end{table}

@@

The \textit{Depression} dataset\cite{bezmaternykhddRestingStateClosed2021,
bezmaternykhBrainNetworksConnectivity2021} includes rs-fMRI scans from
non-depressed controls and mildly- or moderately-depressed subjects. In our analysis,
the binary classification task is \code{depress v control}.

@@

The \code{viglant v nonvigilant} task was formed based on self-report
questionnaire items involving
"vigilance"\cite{gorgolewskiCorrespondenceIndividualDifferences2014}. The
\code{trait\_attend v trait\_nonattend} task was formed using PANAS-X
\cite{watsonPANASXManualPositive1994} items related to self-reported
wakefulness and attention over the past weeks. Finally, the
\code{task\_attend v task\_nonattend} classification task was constructed
using the scores of the Conjunctive Continuous Performance Task
(CCPT\cite{shalevConjunctiveContinuousPerformance2011}). This is a behavioral
performance task which requires the subject to quickly and selectively respond
to only certain visual stimuli. Details allowing exact reproduction of these
splits are available in this paper's
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/preprocess/unify\_attention\_data.py}{source
code}.

@@

We limited preprocessing to a simple pipeline of, in order: (1) brain
extraction, (2) slice timing correction, (3), motion correction, and (4),
registration to the MNI152
\cite{fonovUnbiasedAverageAgeappropriate2011,fonovUnbiasedNonlinearAverage2009}
2mm isotropic, asymmetric template version C made available through
\href{https://www.templateflow.org/}{TemplateFlow}
\cite{ciricTemplateFlowFAIRsharingMultiscale2021}.

@@

We limited preprocessing methods to these steps sometimes because data required
for other typical preprocessing steps (e.g. field intensity maps, physiological
measurements) was missing, but also because all pipeline intermediates were
saved to later compare the effect of increasing degrees of preprocessing on
feature predictive utility (see Section \ref{sec:multiverse}), and it was
important to limit the number of preprocessing steps to prevent excessive
computational costs and comparisons.

@@

\begin{table}[h!]
\caption{
\label{tab:features}
Feature groupings for summarization. eigs = non-RMT eigenfeatures and their combinations with
other non-RMT eigenfeatures. rmt = RMT eigenfeatures and their combinations. tseries = baseline
timeseries reductions.\\
}
\small
\centering
\begin{tabular}{lllr}
\hline
\textbf{Coarse Grouping} & \textbf{Fine Grouping} & \textbf{Feature id}   \\
\hline
eigs           & eigs          & \tcode{eigs}                           \\
               & eigs max      & \tcode{eigsminmax10}                   \\
               &               & \tcode{eigsminmax20}                   \\
               &               & \tcode{eigsminmax5}                    \\
               & eigs middle   & \tcode{eigsmiddle10}                   \\
               &               & \tcode{eigsmiddle20}                   \\
               &               & \tcode{eigsmiddle40}                   \\
               & eigs smooth   & \tcode{eigs + eigs\_smooth}            \\
               &               & \tcode{eigs + savgol}                  \\
               &               & \tcode{eigs\_savgol}                   \\
               &               & \tcode{eigs\_smooth}                   \\
               &               &                                        \\
rmt            & rmt + eigs    & \tcode{eigs + levelvar}                \\
               &               & \tcode{eigs + rigidity}                \\
               &               & \tcode{eigs + rigidity + levelvar}     \\
               &               & \tcode{eigs + unfolded}                \\
               &               & \tcode{eigs + unfolded + levelvar}     \\
               &               & \tcode{eigs + unfolded + rigidity}     \\
               & rmt only      & \tcode{levelvar}                       \\
               &               & \tcode{rigidity}                       \\
               &               & \tcode{rigidity + levelvar}            \\
               &               & \tcode{unfolded}                       \\
               &               & \tcode{unfolded + levelvar}            \\
               &               & \tcode{unfolded + rigidity}            \\
               &               & \tcode{unfolded + rigidity + levelvar} \\
               &               &                                        \\
tseries        & location      & \tcode{T-max}                          \\
               &               & \tcode{T-mean}                         \\
               &               & \tcode{T-med}                          \\
               &               & \tcode{T-min}                          \\
               &               & \tcode{T-p05}                          \\
               &               & \tcode{T-p95}                          \\
               & scale         & \tcode{T-iqr}                          \\
               &               & \tcode{T-rng}                          \\
               &               & \tcode{T-rrng}                         \\
               &               & \tcode{T-std}                          \\
\hline
\end{tabular}
\end{table}

@@

The time and space complexities of computing the full \(N \times N\)
correlation matrix of \(\mathbf{M}\) (and subsequent eigenvalues) are too
large to be tractable. However, since transposition does not change the
eigenvalues, we can use the transpose to efficiently compute these eigenvalues
(see Appendix \ref{sec:transpose}). These raw eigenvalues (feature \code{eigs}
in Table \ref{tab:features}), most directly test the idea that eigenvalues
alone are useful functional connectivity features.

@@

As eigenfeatures, we compute the unfolded eigenvalues, and the spectral
rigidity and level number variance for all \(L \in \{ 1, 2, \dots 20 \}\).
These are \code{unfolded}, \code{rigidity}, and \code{levelvar}, respectively,
in Table \ref{tab:features}.

\textit{Trimming Variants}. As the unfolding procedure operates on the sorted
eigenvalues, and involves fitting a smooth polynomial or exponential to these
values, extreme values are often omitted from fitting
\cite{moralesImprovedUnfoldingDetrending2011,
abueleninEffectUnfoldingSpectral2012}. The number of eigenvalues to trim is
subjective, and unfortunately, with a large number of different source
matrices, it is too destructive to naively use the same hard criterion (e.g.
largest 10\%, largest three) for all unfoldings.

@@

\textit{Unfolding Variants}. Following any trimming, the eigenvalues are fit
with a polynomial. Because the choice of degree has been known to dramatically impact
certain analyses involving the spectral rigidity or level
variance\cite{abul-magdUnfoldingSpectrumChaotic2014,
moralesImprovedUnfoldingDetrending2011, abueleninSpectralUnfoldingChaotic2018,
fossionRandommatrixSpectraTime2013, abueleninEffectUnfoldingSpectral2012}, we
compute the rigidity and level variance with all possible combinations of
trimming and unfolding polynomial degrees of 3, 5, 7, and 9.

\textit{Software Release}. Because the computation of these statistics is non-trivial, and to make the
unfolding procedure more transparent and reproducible, we developed and release
the separate, open-source Python library
\code{empyricalRMT}\cite{dm-bergerStfxecutablesEmpyricalRMTV12022}. The library
allows for efficient, parallel computation of the spectral rigidity and level
variance via Monte Carlo methods, and automatically ensures convergence of the
statistics to user-specified tolerances. The library also makes available
various other functions useful for empirical RMT analyses, such as unfolding
and trimming functions, and plotting facilities for classic RMT ensembles.

@@

When computing RMT eigenfeatures, the polynomial fitting during unfolding has a
smoothing effect, as do the additional averaging \(\langle\;\rangle_c\)
operations. However, it is possible other, simpler transformations of the
eigenvalues (e.g. uniform or Savitsky-Golay smoothing) might have equal or
greater predictive utility. We thus also test smoothed variants of the
eigenvalues as predictive features: the sorted eigenvalues are smoothed with
either a uniform (moving average) filter or Savitsky-Golay, using window sizes
of 3, 5, 7, and 9 to yield 8 total features (\code{eigs\_smooth} and
\code{eigs\_savgol} in Table \ref{tab:features}).

@@

A combination of RMT and non-RMT eigenfeatures could be more predictive than
either alone. We test this with a variety of eigenfeature combinations, with a
focus on combinations that involve the simplest features (e.g. raw eigenvalues,
unfolded eigenvalues) and then features involving additional processing (e.g.
smoothed eigenvalues, level variance, rigidity). Combined features are formed
by concatenation, so that if we have features \(f_1, \dots,  f_n\) with
dimensions \(p_1, \dots, p_n\), then the combined feature is \([f_1; \dots;
f_n]\) with \(\sum p_i\) dimensions.  The final combinations chosen can be
found in Table \ref{tab:features}, where the "\code{+}" symbol indicates concatenation.
